#Introduction
============================
This is the first project of the course Reprodicible Research from the Data Science Specialization.

##Synopsis

The purpose of this project was:
- loading data and processing it
- handling missing values
- interpreting data to answer reasearch questions

##Data

The data for this assignment was downloaded from the course web site:

Dataset: Activity monitoring data [52K]
The variables included in this dataset are:

- steps: Number of steps taking in a 5-minute interval (missing values are coded as NA)

- date: The date on which the measurement was taken in YYYY-MM-DD format

- interval: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

##Loading Data

Downloading the data and loading it into a variable:
```{r}
if(!file.exists("getdata-projectfiles-UCI HAR Dataset.zip")) {
        temp <- tempfile()
        download.file("http://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip",temp)
        unzip(temp)
        unlink(temp)
}

data <- read.csv("activity.csv")
```
##What is the the mean total number of steps taken each day?
Sum steps by day,create Histogram and calcualte mean and median:
```{r,echo=FALSE}
steps_by_day <- aggregate(steps ~ date, data, sum)
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")
rmean <- mean(steps_by_day$steps)
rmedian <- median(steps_by_day$steps)
```
The mean is r rmean and the median is r rmedian
##What is he average of the daily activity pattern?
- Calculate the average steps per interval each day
- Plot the average no. of steps for each day by intervals.
- Find the interval with te highest average no. of steps.
```{r,echo=FALSE}
steps_by_interval <- aggregate(steps ~ interval, data, mean)

plot(steps_by_interval$interval,steps_by_interval$steps, type="l", xlab="Interval", ylab="Number of Steps",main="Average Number of Steps per Day by Interval")

max_interval <- steps_by_interval[which.max(steps_by_interval$steps),1]
```
The 5-minute interval, on average across all the days in the data set, containing the maximum number of steps is `r max_interval`.

##Impute missing values. Compare between imputed and non-imputed data

Missing data needed to be imputed. Only a simple imputation approach was required for this assignment. Missing values were imputed by inserting the average for each interval. Thus, if interval 10 was missing on 10-02-2012, the average for that interval for all days (0.1320755), replaced the NA.
```{r}
incomplete <- sum(!complete.cases(data))
imputed_data <- transform(data, steps = ifelse(is.na(data$steps), steps_by_interval$steps[match(data$interval, steps_by_interval$interval)], data$steps))
```
Zeroes were imputed for 10-01-2012 because it was the first day and would have been over 9,000 steps higher than the following day, which had only 126 steps. NAs then were assumed to be zeros to fit the rising trend of the data.
```{r}
imputed_data[as.character(imputed_data$date) == "2012-10-01", 1] <- 0
```

Recount total steps and create a histogram

```{r}
steps_by_day_i <- aggregate(steps ~ date, imputed_data, sum)
hist(steps_by_day_i$steps, main = paste("Total Steps Each Day"), col="blue", xlab="Number of Steps")

#Create Histogram to show difference. 
hist(steps_by_day$steps, main = paste("Total Steps Each Day"), col="red", xlab="Number of Steps", add=T)
legend("topright", c("Imputed", "Non-imputed"), col=c("blue", "red"), lwd=10)
```

Calculate new mean and median for imputed data.
```{r,echo=FALSE}
rmean.i <- mean(steps_by_day_i$steps)
rmedian.i <- median(steps_by_day_i$steps)
```

Calculate the difference between imputed and non-impputed data.
```{r,echo=FALSE}
mean_diff <- rmean.i - rmean
med_diff <- rmedian.i - rmedian
```
Calculate the total difference
```{r,echo=FALSE}
total_diff <- sum(steps_by_day_i$steps) - sum(steps_by_day$steps)
```

- The imputed data mean is `r rmean.i`
- The imputed data median is `r rmedian.i`
- The difference between the non-imputed mean and imputed mean is `r mean_diff`
- The difference between the non-imputed mean and imputed mean is `r med_diff`
- The difference between total number of steps between imputed and non-imputed data is `r total_diff`. Therefore there a total of `r total_diff` more steps.

##Difference in activity pattern on weekdays and weekends?

Created a plot to compare and contrast number of steps between the week and weekend. There is a higher peak earlier on weekdays, and more overall activity on weekends.
```{r}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", 
              "Friday")
imputed_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imputed_data$date)),weekdays), "Weekday", "Weekend"))

steps_by_interval_i <- aggregate(steps ~ interval + dow, imputed_data, mean)

library(lattice)

xyplot(steps_by_interval_i$steps ~ steps_by_interval_i$interval|steps_by_interval_i$dow, main="Average Steps per Day by Interval",xlab="Interval", ylab="Steps",layout=c(1,2), type="l")
```
